Eagle: Making multiple-locus association mapping on a genome-wide scale routine

### 2018-07-09 AM+draft.tex
9July refining text v1

\title{Eagle: Making multiple-locus association mapping on a genome-wide scale routine}
\author{Andrew W. George$^1$, Arunas Verbyla$^1$, Joshua Bowden$^2$, need to sort out the other authors}

@@

\section{Simulation Study}

To explore the computational and statistical performance of Eagle, we conducted a large simulation data.  Genome-wide data 
were generated under five different scenarios. These scenarios were data where the sample size was X and the number of snp Y (X x Y) ,

We generated data under five different scenarios.

We generated 
replicates via data perturbation (ref: Zhao et al. 2007).

These are a GWAS of size 150 individuals and 5000 snp (150 X 5K), 350 individuals and 400000 snp (350 X 400K),  1500 individuals and 
50000 snp (1500 x 50K), 2000 individuals and 500000 snp (2000 x 500K), 4000 individuals and 
1500000 snp (4000 x 1.5M), and 10000 individuals and 1500000 snp (10000 x 1.5M).  
We chose these scenarios to reflect some of the different sized GWAS being performed in animals, plants, and humans. 
For each scenario, we generated 100 replicates of data. A single replicate consists of snp genotype data and quantitative trait data. 
We obtained the snp data from the publicly available 1000 genome project (phase 3). The quantitative trait data we generated from the 
snp data by selecting a set of snp loci, assigning allelic effects to these snp, and aggregating these effects for each individual along with 
random error. The number of snp selected per replicate follows a Poisson distribution with mean 30.
 The quantitative trait was generated to have a heritability of 50\%.
Analyses were performed on a high end desktop computer. It had dual 8-core Xeon processors, three Kepler Tesla GPUs, and 128 Giggabytes of RAM. All implementations except, GEMMA, made use of distributed computing, either explicitly or implicitly through multi-threaded BLAS/LAPACK libraries.

Replicates are generated with data perturbation (Zhao et al. 2007). Data perturbation makes use of the phenotypic and genotypic data observed in a real study to create a simu- lated (quantitative) trait. It affords us the opportunity to generate replicates whose analysis more closely mirrors the complexities of analyzing real data. For computational expediency, our simulation study is based on data observed from the Biloela site in 2006. In generating trait data for a replicate, we assume a broad based heritability of 0.6, there is a single major QTL, the polygenic component consists of 30 polygenes where the effect of a polygene is formed from an unmapped SNP which was chosen randomly, and the variance structure of the simulated trait closely follows that of the loaf volume trait. The marker data consists of those genotypes collected in the association study from the 3129 SNP across the 21 homologous chromosomes. Each replicate has the same marker data. It is the simulated trait data that varies across replicates.
Our simulation study consists of three parts. First, we conduct a null study to investigate whether our Monte Carlo sampling approach controls correctly the genome- wide type I error rate. We also evaluate the impact of assuming a block diagonal structure for V. We do this by
also calculating the genome-wide type I error rates when the variance matrix for the joint distribution of the test sta- tistic is formed from loci pairings across the entire genome. Ten thousand replicates are generated where the QTL has no effect. Results are reported for QK-based association mapping implemented as a single-stage and weighted two- stage analysis.
Second, a power study is performed to investigate if there is a difference in performance between implementing QK-based association mapping as a single-stage analysis versus a two-stage analysis. Data are generated under QTL of different sizes and with varying percentages of missing marker data (1.1, 5, 9, 11 %). Missing marker genotypes are imputed for a marker locus by coding its SNP geno- types as ?1 and 1, normalizing its marker data, and replac- ing missing genotypes with zero values. One thousand replicates are generated for each of the four different per- centages of missing marker data and each differently sized QTL. A genome-wide signi cance level of 5 % is assumed.
Third, a study to evaluate our heuristic procedure for choosing between single-stage and two-stage analysis is performed. Data are generated under a QTL of no, mod- erate, and large effect. One thousand replicates are gen- erated for each differently sized QTL. The performance of our heuristic procedure is measured by calculating the proportion of replicates for which the analysis type is inferred correctly. To determine if a single-stage analy- sis is truly needed, it is necessary to analyze fully the genome-wide data via single-stage and weighted two-stage analysis. Then, for each marker locus j, perform the test
| ? log padj ? log qadj| > | ? log padj ? 1.3|, where we have jjj
assumed a genome-wide signi cance level of 5 %. If true, an incorrect  nding occurs if our heuristic procedure deter- mines that it is valid to implement two-stage analysis.
\section{Implementation}

\section{New section written on iPhone added 5 Feb 2018}
\section{Generation of Simulation Data}

The data were generated via data perturbation. Data perturbation amalgamates real with simulated data to generate replicates. 
It is a way of introducing greater realism into a simulation study. 
 Here, the genotype data are real, the 
quantitative trait data are simulated. 
 The SNP genotypes were drawn, 
according to the specifications of the scenario, from data collected from the 1000 Genome Project, version 3   \cite{10002010map}. 
Across scenarios, the SNP data differs. Across replicates within a scenario, the SNP data are the same.

To generate the trait data $y$, first, $q$, the number of SNP that are to be assigned a quantitative value is drawn from a Poisson distribution with 
mean 30. Second, $q$ SNP are selected randomly. Third, we assume an additive model for the SNP. The SNP genotypes AA, AB, and BB 
are assigned the values -1, 0, and 1, respectively. Fourth, the SNP effects are summed across the $q$ selected loci, for each individual, to 
generate a $g^{(n \times 1)}$ vector of genetic values where $n$ is the number of individuals. 
Fifth, $e^{(n \times 1)}, a vector of residuals, is drawn from a normal distribution where $e_i \sim N(0, \sigma^2_e)$ and $\sigma^2_e$ is 
the residual variance that has been set to yield a trait with heritability 0.5. Sixth,  the trait data are formed as $y =  g + e$.

>>>>

\title{Eagle: Making multiple-locus association mapping on a genome-wide scale routine}
\author{Andrew W. George$^1$, Arunas Verbyla$^1$, Joshua Bowden$^2$}

@@

\section{Generation of Simulation Data}

The data were generated via data perturbation (Zhao et al. 2007). Data perturbation amalgamates real with simulated data to generate replicates. 
It is a way of introducing greater realism into a simulation study. 
 Here, the genotype data are real, the 
quantitative trait data are simulated. 
 The SNP genotypes were drawn, 
according to the specifications of the scenario, from data collected from the 1000 Genome Project, version 3   \cite{10002010map}. 
Across scenarios, the SNP data differs. Across replicates within a scenario, the SNP data are the same.

To generate the trait data $y$, first, $q$, the number of SNP that are to be assigned a quantitative value is drawn from a Poisson distribution with 
mean 30. Second, $q$ SNP are selected randomly. Third, we assume an additive model for the SNP. The SNP genotypes AA, AB, and BB 
are assigned the values -1, 0, and 1, respectively. Fourth, the SNP effects are summed across the $q$ selected loci, for each individual, to 
generate a $g^{(n \times 1)}$ vector of genetic values where $n$ is the number of individuals. 
Fifth, $e^{(n \times 1)}$, a vector of residuals, is drawn from a normal distribution where $e_i \sim N(0, \sigma^2_e)$ and $\sigma^2_e$ is 
the residual variance that has been set to yield a trait with heritability 0.5. Sixth,  the trait data are formed as $y =  g + e$.
